\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}

### 2014-06-19 paper/semantics.tex
major reorganizing/rewriting of section 3.

Although there is a strict one-way dependence between the two stages at the type
level, their terms are defined mutually recursively:
$\next$ embeds stage-\bbtwo{} expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
These $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with \{braces\} in our concrete syntax to clearly indicate the stage boundaries with in a program.
The problem of how to turn an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.

The grammar and type system for our language, \lang, are given in
\ref{fig:grammar} and \ref{fig:statics}.
We formulated our typing judgments in the style of \cite{davies96}, where the whole judgment is annotated with a stage.  
Variables in the context are annotated with a stage and must be used at the same stage at which they were declared.
All of the non-stage constructs have the expected type-level behavior.  
This is made manifest as rules which are entirely abstract over stage.

In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, $\next$, given an argument with type $A$ at stage \bbtwo, forms a $\fut A$ at stage \bbone.  
That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
This promise can only be redeemed be the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.

@@

\subsection {Staged Fast Exponentiation}

@@

Without going into details of how recursion is handled, we implement staged fast exponentiation in \lang~below.  
We use {\tt \$} in place of $\fut$ in this an other examples.

@@

Although the code for {\tt fexp} looks very much like the stage-less mathematical definition above, it is in fact a staged program.  
See how the {\tt if}s and exponent decomposition are all at stage \bbone, since they occur within $\prev$ blocks.  
Essentially, the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the stage \bbone~without requiring us to disrupt the elegant functional structure of the code.

This example also illustrates how stage \bbone~and stage \bbtwo~code can be arbitrarily nested.  Normal notions of eager evaluation reduce code in a leaf-to-root way, but this conflicts with the precept that we evaluate stage \bbone~before stage \bbtwo.  
To resolve this conflict, we need a formal semantics for \lang.

\subsection{Semantics}

@@

%Previous work (\cite{davies96}) focuses on the correspondence between the type system and existing temporal logics, whereas we care more about operational behavior and cost.  In this section, we'll consider a few proposals for our language before settling on one we like.  All of the proposals are call-by-value, differing primarily in how they handle values of $\fut$ type.

@@

In this section, we discuss the dynamic semantics of our language.  
Before we can get into the details of the semantics, we have to first make some decisions about the behavior we want it to have.  
Similar to \cite{davies96}, we want our semantics to have the form of a judgment that partially evaluates an expression to create a residual that can be evaluated at the second stage in the standard way.  Again like Davies, we choose a call-by-value paradigm as our baseline at both stages.  The only significant freedom that this leaves is how to handle the staging constructs.

We present here two reasonable interpretations of these staging constructs, 
in the context of the following stage-\bbtwo~example:
\begin{lstlisting} 
prev{
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
The first option is to duplicate the contents of the first $\next$ expression to produce,
\begin{lstlisting}
(4+5) * (4+5) * 15
\end{lstlisting}
Essentially, we're treating $\next$ as a value a stage-\bbone, so long as all of its contained $\prev$s are reduced.  
This is the technique of \cite{davies96}.
Instead of this, we take a second, nearly equivalent approach.  
That is to bind the contents of the first $\next$ expression to some variable (here, $x1$) and then duplicate that variable,
\begin{lstlisting} 
let x1 = 4+5 in
x1 * x1 * 15
\end{lstlisting}

We think that this interpretation makes more sense from the point of view of cost, 
since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
This is {\em nearly} equivalent to the first expression, because they have different termination behavior 
when a non-terminating stage-\bbtwo~expression is declared but never used.

Now that we know {\em what} the behavior is that we want, we can explain {\em how} our semantics achieves that behavior.  We start with the same example:
\begin{lstlisting}
prev {
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
The first thing we need to do is evaluate the binding for $x$.  Of course, we can't evaluate inside the $\next$ statement, but we still need to produce a value that will get substituted.  What we do instead is replace the body of the $\next$ by a variable and make an explicit substitution, which then floats up to the top of the containing $\prev$:
\begin{lstlisting} 
prev{
[4+5/x1]
let x = (next{x1}, 15) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
prev{
[4+5/x1]
next{
	prev{#1 (next{x1}, 15)} 
	* prev{#1 (next{x1}, 15)} 
	* prev{hold (#2 (next{x1}, 15))}
}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
prev{
[4+5/x1]
	next{ x1 * x1 * 15 }
}
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 4+5 in
 x1 * x1 * 15
\end{lstlisting}

Once again, the point here is that we did not duplicate the $4+5$ in our residual.  
Instead we lifted it out into a substitution, leaving a hatted variable [note: add hats above] in its place,
and we reified the substitution into a let-binding at the enclosing $\prev$.

\subsection{more formally...}

Abstractly, we can think of our evaluation as proceeding in the standard way for stage-1 code. When the evaluator encounters a $\next \{e\}$ expression, it places $e$ off to the side in a context and replaces the whole expression with a reference to the context entry.  These references are then passed around as stage-1 values for $\fut$ types.  But what if $e$ contains $\prev$ expressions?  To ensure that all stage-1 code is evaluated before any stage-2 code, we must evaluate all of the 1-code contained in $e$ before inserting it into the table.  This entails searching $e$ for all contained $\prev$s and evaluating them in place.

More formally, evaluation comprises the following judgments:

\begin{center}
\begin{tabular}{|c|p{2.5cm}|p{2.5cm}|} \hline
Judgment & Reads & Conditions \\ \hline 
$\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
$\diaonesub$ & ``$e$ evaluates to future table $\xi$ and value $v$'' 
& $\typesone e A$ 
	\newline $\isvalone v$ 
	\newline ... \\ \hline 
$\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
...&...&... \\ \hline
\end{tabular}
\end{center}

Evaluation and speculation are the main judgments here, the rest being largely administrative.  The evaluation judgment operates on stage-1 code, whereas the speculation operates on stage-2 code.  Since $\next$ and $\prev$ are the crossover points between 1-code and 2-code, they are correspondingly the only places where the evaluation and speculation judgments depend on the other.

% FIXME we need a real name for hatted variables
The evaluation judgment is very similar to standard call-by-value evaluation, as goal 1 would suggest.  The input to evaluation is a stage-1 expression (usually $e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered later.  Evaluation has two outputs: the {\em future context} (usually $\xi$) and the {\em partial value} (usually $v$).  We cover the latter first.  The partial value is essentially the result of the first-stage portion of $e$, and must be a \bbone-value of the same type as $e$.  As usual, this means that $v$ must be composed only of base primitives, tuples, injections, and lambdas (corresponding to base types, products, sums, and functions).  Analogously, the value corresponding to $\fut$ is a construct called a {\em hatted variable} (written $\hat y$), which signifies a reference to some stage-2 computation.  The mapping between these hatted variables and stage-two expressions is

...

\subsection{Other nice properties}

\begin{enumerate}
\item For some $\colone{e}{A}$ containing no $\next$ expressions, the semantics is derivationally equivalent to standard call-by-value evaluation.
\end{enumerate}

@@

\TODO unfinished; where should this go? combine with table above

\noindent\TODO terminology for $\Xi,\Gamma,\xi,v,q$?

\noindent\TODO need to define 1-vals at 1 and 2, and 2-vals at 2. what is the
terminology? (?, residuals, results?)

\noindent\TODO put $\Xi$ and $\Gamma$ in the judgments where appropriate

\noindent\TODO be uniform about which judgments have which contexts, etc

@@

\noindent\TODO don't call it a table

\noindent\TODO what is the notation for $\xi$s? how do you build them?

\noindent\TODO define reification explicitly as a metafunction

\begin{definition}
A table $\xi$ is well-formed ($\xi\wf$) if either:
\begin{enumerate}
\item $\xi = \cdot$; or
\item $\xi = \xi',x\mapsto e$ where
$\Xi,\dom{\xi'};\Gamma\vdash e : B @ \bbtwo$ and
$\Xi,\dom{\xi'};\Gamma\vdash e \valot$.
\end{enumerate}
\end{definition}

>>>>

Although there is a strict one-way dependence between the two stages at the type
level, their terms are defined mutually recursively:
$\next$ embeds stage-\bbtwo{} expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
These $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with \{braces\} in the concrete syntax to clearly indicate the stage boundaries with in a program.
The problem of how to turn an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.

%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.

The grammar and type system for our language, \lang, are given in
\ref{fig:grammar,fig:statics}.
We annotate the typing judgment and context variables with stages; only $\fut$
and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.

Specifically, $\next$, given an argument with type $A$ at stage \bbtwo, forms a $\fut A$ at stage \bbone.  
That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
This promise can only be redeemed be the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.

@@

\subsection{Staged Evaluation}

@@

Ignoring for the moment how to define recursive functions in \lang, we can
implement staged fast exponentiation as follows:%
\footnote{We render $\fut$ as {\tt \$} in code examples.}

@@

Although the code for {\tt fexp} looks very much like the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if}s and exponent
decomposition are all at stage \bbone, since they occur within $\prev$ blocks.
Essentially, the type system is sufficient to prove the observation that all of
the decomposition of the exponent can be done at the stage \bbone~without
requiring us to disrupt the elegant functional structure of the code.

This example also illustrates how stage \bbone{} and stage \bbtwo{} code can be
arbitrarily nested.  Ordinary evaluation always eliminates outermost redexes
first, but this conflicts with the precept that we evaluate all stage \bbone{}
code before stage \bbtwo{} code, in the case that stage \bbone{} code is inside
stage \bbtwo{} code.

A suitably staged dynamic semantics for \lang{} must therefore evaluate
\emph{all} the stage \bbone{} subexpressions of a term before its stage
\bbtwo{} subexpressions.

Consider how we could achieve this in the following stage \bbtwo{} term:
\begin{lstlisting} 
prev{
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
One option is to duplicate the contents of the first $\next$ expression, to produce
\begin{lstlisting}
(4+5) * (4+5) * 15
\end{lstlisting}

Here we are essentially treating $\next\{e\}$ as a stage \bbone{} value, so long
as all the $\prev$ subterms of $e$ are reduced. This is the approach taken by
\cite{davies96}.

We take a different approach in our dynamic semantics, which is to let-bind the
contents of the first $\next$ expression to some variable (here, $x1$) and
duplicate that variable,
\begin{lstlisting} 
let x1 = 4+5 in
x1 * x1 * 15
\end{lstlisting}

We think that this interpretation makes more sense from the point of view of cost, 
since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
This is \emph{nearly} equivalent to the first expression, because they have different termination behavior 
when a non-terminating stage-\bbtwo~expression is declared but never used.

How do our semantics achieve this behavior? In the original term,
\begin{lstlisting}
prev {
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
we must start by evaluating the binding for $x$.  We can't evaluate inside the
$\next$ statement, but must still substitute some result into the body of the
\texttt{let}. We replace the contents of $\next$ with a variable and make an
explicit substitution, which then floats up to the top of the containing
$\prev$:
\begin{lstlisting} 
prev{
[4+5/x1]
let x = (next{x1}, 15) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
prev{
[4+5/x1]
next{
	prev{#1 (next{x1}, 15)} 
	* prev{#1 (next{x1}, 15)} 
	* prev{hold (#2 (next{x1}, 15))}
}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
prev{
[4+5/x1]
	next{ x1 * x1 * 15 }
}
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 4+5 in
 x1 * x1 * 15
\end{lstlisting}

Once again, the point here is that we did not duplicate the $4+5$ in our residual.  
Instead we lifted it out into a substitution, leaving a hatted variable [\TODO add hats above] in its place,
and we reified the substitution into a let-binding at the enclosing $\prev$.

\subsection{Dynamics}

@@

%{{{ stuff nico commented out
%Previous work (\cite{davies96}) focuses on the correspondence between the type system and existing temporal logics, whereas we care more about operational behavior and cost.  In this section, we'll consider a few proposals for our language before settling on one we like.  All of the proposals are call-by-value, differing primarily in how they handle values of $\fut$ type.

@@

%}}}
%{{{ stuff carlo commented out
%Abstractly, we can think of our evaluation as proceeding in the standard way for
%stage-1 code. When the evaluator encounters a $\next \{e\}$ expression, it
%places $e$ off to the side in a context and replaces the whole expression with a
%reference to the context entry.  These references are then passed around as
%stage-1 values for $\fut$ types.  But what if $e$ contains $\prev$ expressions?
%To ensure that all stage-1 code is evaluated before any stage-2 code, we must
%evaluate all of the 1-code contained in $e$ before inserting it into the table.
%This entails searching $e$ for all contained $\prev$s and evaluating them in
%place.

%The evaluation judgment operates on stage-1 code, whereas the
%speculation operates on stage-2 code.  Since $\next$ and $\prev$ are the
%crossover points between 1-code and 2-code, they are correspondingly the only
%places where the evaluation and speculation judgments depend on the other.

%The evaluation judgment is very similar to standard call-by-value evaluation.
%The input to evaluation is a stage-1 expression (usually
%$e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered
%later.  Evaluation has two outputs: the {\em future context} (usually $\xi$) and
%the {\em partial value} (usually $v$).  We cover the latter first.  The partial
%value is essentially the result of the first-stage portion of $e$, and must be a
%\bbone-value of the same type as $e$.  As usual, this means that $v$ must be
%composed only of base primitives, tuples, injections, and lambdas (corresponding
%to base types, products, sums, and functions).  Analogously, the value
%corresponding to $\fut$ is a construct called a {\em hatted variable} (written
%$\hat y$), which signifies a reference to some stage-2 computation.  The mapping
%between these hatted variables and stage-two expressions is
%}}}

% TODO move these to macros.tex once they're finalized
\newcommand{\pval}{\ensuremath{\mathsf{pval}}}
\newcommand{\res}{\ensuremath{\mathsf{res}}}
\renewcommand{\val}{\ensuremath{\mathsf{val}}}

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item Partial values ($\pval$s) are stage \bbone{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|next {x1 * x1 * 15}|.
\item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|4+5|.
\item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|1215|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone{} term to a pair of an
environment $\xi$ mapping newly-created variables to stage \bbtwo{} residuals,
and a partial value $v$ which may refer to those variables. This judgment
necessarily traverse stage \bbtwo{} subexpressions in order to find their stage
\bbone{} subexpressions; we call this process \emph{speculation}. Speculation is
implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo{} term to a
residual.

\TODO finish explaining how the nested evaluation actually works.
explain reification, hatted variables, and contexts.

A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
given a type of the form $A\to\fut(B\to C)$.%
\footnote{We can rewrite \texttt{fexp} in this form, or simply apply the following
higher-order function which makes the adjustment:

\TODO write the function here}
Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
function:
$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in \langTwo.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.

\TODO make sure this is true (haven't written the canonical forms theorems yet)

That this sequence of evaluations is in fact staged follows from our
characterizations of partial values, residuals, and values, that $\redonesym$
outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.

\begin{remark}
For any $\colone{e}{A}$ containing no $\next$ subexpressions, $\redonesym$ will
always compute an empty environment, and a partial value identical to the result
of call-by-value evaluation of $e$.
%derivationally equivalent to standard call-by-value evaluation.
\end{remark}

\TODO 
If we change the highlighted rules to something slightly different, then we get
precisely the semantics from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.

@@

\noindent\TODO terminology for the contexts?

@@

\begin{definition}
An environment $\xi$ is well-formed ($\xi\wf$) if either:
\begin{enumerate}
\item $\xi = \cdot$; or
\item $\xi = \xi',x\mapsto e$ where
$\Xi,\dom{\xi'};\Gamma\vdash e : B @ \bbtwo$ and
$\Xi,\dom{\xi'};\Gamma\vdash e \valot$.
\end{enumerate}
\end{definition}

### 2014-06-19 paper/staging.tex
major reorganizing/rewriting of section 3.

The grammar for types, terms, values, and contexts in \ellStaged are given in Figure \ref{fig:coreSyntax}.  The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.  It would be tempting to merge the two stages into one syntactic class and maintain the stage invariants with the typing judgements.  Such a trick would reduce the total number of rules presented later, but at the cost of making those rules type derivation-directed, rather than syntax directed.  Alternatively, we chose to have one context merged accross both stages, rather than one context for each stage.  The intended effect of this decision is for bindings in one stage to shadow binding to the same variables in the other stage.  The reasons for this will become clear later.

@@

The static semantics for \ellStaged~is given in Figure \ref{fig:coreStatics}.  Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).  The static semantics comprises two judgements.  Those are:

>>>>

The grammar for types, terms, values, and contexts in \ellStaged are given in \ref{fig:coreSyntax}.  The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.  It would be tempting to merge the two stages into one syntactic class and maintain the stage invariants with the typing judgements.  Such a trick would reduce the total number of rules presented later, but at the cost of making those rules type derivation-directed, rather than syntax directed.  Alternatively, we chose to have one context merged accross both stages, rather than one context for each stage.  The intended effect of this decision is for bindings in one stage to shadow binding to the same variables in the other stage.  The reasons for this will become clear later.

@@

The static semantics for \ellStaged~is given in \ref{fig:coreStatics}.  Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).  The static semantics comprises two judgements.  Those are:
